Is this the longest Chess game?

### 2020-03-15 trunk/chess/papers/longest.tex
Almost done with this boring paper

@@

\title{Is this the longest Chess game?}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2020 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2015 with the {\tt ShortMoveString} of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
% XXX should match the length of the game
\textcurrency 17,697
}
}

@@

Many chess moves are reversible (e.g. moving the knight forward and
back to its starting position~\cite{survival}), so informal games of
chess could last forever with the players repeating a short cycle. In
AD~1561, Ruy L\'opez added the ``fifty-move rule'' to prevent infinite
games.\footnote{This rule only applied to games started after its
  introduction, so it is possible that some pre-1561 games are still
  in progress and may never end.} This rule (detailed below) ensures
that irreversible moves are regularly played, and so the game always
makes progress towards an end state. Another rule, ``threefold
repetition'' also guarantees termination as a sort of backup plan
(either of these rules would suffice on its own).

% This is good for the insurers and reinsurers of chess tournaments
% (games are guaranteed to end, rather than simply ending with
% probability 1).
So, chess is formally a finite game. This good for computer
scientists, since it means that chess has a trivial O(1) optimal
solution. This allows us to move onto other important questions, like:
What is the longest chess game? In this paper I show how to compute
such a game, and then gratuitously present all of its
% XXX check that this is the final one
17,697 
%
half-moves. Even if you are a chess expert (``chexpert''), I bet you
will be surprised at some of the corner cases in the rules that are
involved.

Speaking of rules, let's first detail the two main rules that limit the
length of the game.

@@

Draws after 75 moves (per player, so really 150 half-moves) are
compulsory.

@@

% (I seem to remember that this rule used to be about moves, not positions,
% but someone discovered infinite sequences of moves that don't repeat, right?
% I can't find this citation. Maybe can recreat those sequences.)
%
% Fascinatingly (at least as fascinating as anything in this
% questionable undertaking), this rule

@@

The informal version of this rule (``insufficient material'') states
that if neither side has enough pieces to mate the opponent (for
example, a king and bishop can never mate a bare king) then the game
is draw. Again, the formal rule is more subtle:

\begin{quote}
5.2.2. The game is drawn when a position has arisen in which neither
player can checkmate the opponent's king with any series of legal
moves. The game is said to end in a `dead position'. This immediately
ends the game \ldots
\end{quote}

@@

The insufficient material rule is curious in that it requires
nontrivial computation to implement. In order to know whether the
position is a draw, an implementation needs to be able to decide
whether or not a series of moves that results in checkmate exists.
Note that this is not nearly as bad as normal game tree search because
the two sides can collaborate to produce the mate (it is not a
$\exists \forall \exists \forall \ldots$ but rather $\exists \exists
\exists \exists \ldots$). Still, such ``helpmates'' can still be
quite deep (dozens of moves) and are interesting enough to be a
common source of chess puzzles. Proving the non-existence of a
helpmate can be very difficult indeed (Figure~\ref{fig:nohelp}).

@@

% XXX might be a good place to talk about the self-referential "maximal"
% and "minimal" interpretations of this rule?
% Need to have a  \label{sec:ambiguity} somewhere.

@@

It is generally not hard to avoid repeating positions, so the main
obstacle we'll face is the 75-move rule. Let's call an irreversible
move that resets the 75-move counter a {\em critical move}; this is
a pawn move or capturing move (or both). The structure of the game
will be a series of critical moves (I will call these ``critical''
moves) with a maximal sequence of pointless reversible moves in
between them. If we execute the maximum number of critical moves and
make 149 half-moves (just shy of triggering the compulsory version of
the 75-move rule) between them, then this will be fairly easily
seen as a maximal game.

In fact, for most positions, it is easy to waste 149 moves and return
to the exact same position. So, the strategy for generating the
longest game can mostly be broken into two phases: Make a game with a
maximum number of critical moves (it can also contain other moves) and
then pad that game out to maximum-length excursions in between its
critical moves.

\subsection{Maximal critical moves}

Critical moves are pawn moves and captures. There are 16 pawns, each
of which has 6 squares to move into before promoting, so this is
$16\times 6$ critical moves. There are 14 capturable pieces, plus the
16 pawns (they can be captured after promoting) so this nets an
additional $16+14$ critical moves for a total of $126$. Each critical
move can be made after a maximum of $75+74$ reversible moves, giving
$(75+74+1) \times 126 = 18900$ moves. The final move would capture the
last piece, yielding a draw due to the remaining kings being insufficient
material (``dead position''). This is the upper bound.

@@

Each of these captures is {\em both} a pawn move {\em and} a capturing
move, however. This means that we lose $4+4$ critical moves off our
total budget. $150 \times (126 - 8) = 17700$ is the new upper bound.

@@

Note that white is quite constrained during this beginning phase, as
pawn moves are critical moves and must be avoided. Only the white
knights can escape the back rank. When we try to insert $149$
pointless moves, we'll only be able to move the knights and rooks, and
doing this $75$ times must leave e.g. one of the knights on an
opposite-colored square.\footnote{Knight moves always change the color
  of the knight's square, and same for rook moves constrained to the
  a1/b1 or g1/h1 squares.} So we have to be a little careful about the
position in which we make black's first critical move.

Since white can only free their two knights, these are the only pieces
that can be captured by black pawns. So it will not be possible for
black to double four sets of their pawns as in
Figure~\ref{fig:doubled}. This would require white to have a phase of
critical moves to free moves to capture, then black again to finish
doubling pawns, then white again to promote its freed pawns. Each switch
costs move off the theoretical max. We can be more efficient with an
asymmetric approach.

@@

\comment{
1. Nf3 b6 2. Nc3 g6 3. Ne5 g5 4. Nd5 b5
5. Nc4 bxc4 6. Nf4 gxf4 7. Rg1 c6 8. Rh1 c5
9. Rg1 f6 10. Rh1 f5 
}
\begin{center}
\chessboard[setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11]
\end{center}

The white b and g pawns have a clear route to promotion. White can
free each of the remaining 6 pawns with a single capture. Between this
and black's two pawn captures, this is the minimum of 8 pawn moves
that are also captures. Since black has plenty of freed pieces, white
can promote all of their pawns during their own phase of critical
moves, resulting in:

@@

(this is three switches)

\paragraph{Ending condition.} The way this game ends is subtle for
several reasons. First, note that we left the white queen on the board
and used it for mate. It is required that white be the one mating for
parity reasons, similar to the reason black must make the first critical
move. Since white makes critical moves in the last phase, black leads
on non-critical moves; at the moment white makes the checkmating move,
black has made 75 non-critical moves, and white 74. White's 75th move
mates.

@@

Why not capture the white queen too? This would be a critical move,
but once two kings remain, the game ends immediately in a draw. So
there is no length advantage here over checkmate. In fact we could do
worse: Consider the position in Figure~\ref{fig:insufficient} where
black is {\em forced} to capture the queen. This game is over prior to
the capture, so this is shorter than the checkmating sequence! Even if
the king had an escape square, it is arguable
(Section~\ref{sec:ambiguity}) that the game is over in any case: There
is no legal sequence of moves that leads to mate. Either the king
captures the queen (and then clearly no mate is possible with just the
two kings) or the king escapes, but in doing so plays the
75\th\ non-critical move, and triggers a draw by the 75-move rule.

@@

TODO

\section{Future work}

Some puzzles?

\section{A longest game}

@@

(a long game goes here)

Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do
eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad
minim veniam, quis nostrud exercitation ullamco laboris nisi ut
aliquip ex ea commodo consequat. Duis aute irure dolor in
reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla
pariatur. Excepteur sint occaecat cupidatat non proident, sunt in
culpa qui officia deserunt mollit anim id est laborum.

>>>>

\usepackage{adjustbox}

@@

\title{Is this the longest Chess game?}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2020 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2015 with the {\tt ShortMoveString} of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\textcurrency 17,697
}
}

@@

Many chess moves are reversible (e.g.~moving the knight forward and
back to its starting position~\cite{survival}), so informal games of
chess could last forever with the players repeating a short cycle. In
AD~1561, Ruy L\'opez added the ``fifty-move rule'' to prevent infinite
games.\footnote{This rule only applied to games started after its
  introduction, so it is possible that some pre-1561 games are still
  in progress and may never end.} This rule (detailed below) ensures
that irreversible moves are regularly played, and so the game always
makes progress towards an end state. Another rule, ``threefold
repetition'' also guarantees termination as a sort of backup plan
(either of these rules would suffice on its own).

% This is good for the insurers and reinsurers of chess tournaments
% (games are guaranteed to end, rather than simply ending with
% probability 1).
So, chess is formally a finite game. This good for computer
scientists, since it means that chess has a trivial O(1) optimal
solution. This allows us to move onto other important questions, like:
What is the longest chess game? In this paper I show how to compute
such a game, and then gratuitously present all of its 17,697 
moves. Even if you are a chess expert (``chexpert''), I bet you
will be surprised at some of the corner cases in the rules that are
involved.

Speaking of rules, let's first detail the three main rules that limit
the length of the game. These rules cause the game to end in a draw
(tie) when certain conditions are met.

@@

Draws after 75 moves (per player, so really 150 moves) are compulsory.

@@

% (I seem to remember that this rule used to be about moves, not positions,
% but someone discovered infinite sequences of moves that don't repeat, right?
% I can't find this citation. Maybe can recreate those sequences.)
%
% Fascinatingly (at least as fascinating as anything in this
% questionable undertaking), this rule

@@

The informal version of this rule (``insufficient material'') states
that if neither side has enough pieces to mate the opponent (for
example, a king and bishop can never mate a bare king) then the game
is drawn. Again, the formal rule is more subtle:

\begin{quote}
5.2.2. The game is drawn when a position has arisen in which neither
player can checkmate the opponent's king with any series of legal
moves. The game is said to end in a `dead position'. This immediately
ends the game\ldots
\end{quote}

@@

The insufficient material rule is curious in that it requires
nontrivial computation to implement. In order to know whether the
position is a draw, an implementation needs to be able to decide
whether or not a series of moves that results in checkmate exists.
Note that this is not nearly as bad as normal game tree search because
the two sides can collaborate to produce the mate (it is not a
$\exists \forall \exists \forall \ldots$ but rather $\exists \exists
\exists \exists \ldots$). Still, such ``helpmates'' can still be quite
deep (dozens of moves) and are interesting enough to be a common
source of chess puzzles. Proving the non-existence of a helpmate can
be very difficult indeed (Figure~\ref{fig:nohelp}).\footnote{Perhaps
  an enterprising reader can prove that for generalized chess, it is
  NP-hard to decide whether the game has ended due to this rule?}

@@

\subsubsection{Ambiguity} \label{sec:ambiguity}

Moreover, this rule contains some ambiguity. The phrase ``any series
of legal moves'' is usually taken to mean something like, ``the
players alternate legal moves and follow most of the normal rules of
chess.'' In my opinion it is hard to justify an interpretation like
this.

First, the rules specifically define ``legal move'' in Article~3, and
3.10.1 says ``A move is {\em legal} when all the relevant requirements
of Articles 3.1 -- 3.9 have been fulfilled.'' These requirements
describe the movement of each piece as you are familiar (e.g. 3.3
``The rook may move to any square along the file or the rank on which
it stands.''). They also disallow capturing one's own pieces, or
moving when in check. However, they allow as legal some moves that
would otherwise be prohibited, like capturing the opponent's king
(this is excluded by 1.4.1, outside the definition of ``legal'').
Capturing the opponent's king is generally not useful for
demonstrating that checkmate is possible
(Figure~\ref{fig:kingcapture}), so this is mostly a curiosity.

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \chessboard[smallboard,marginleft=false,marginright=false,setfen=BBKBRRnk/PPPPPPnn/3P1BNR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] &
      \chessboard[smallboard,marginright=false,setfen=BBKBRR1B/PPPPnP1n/3P2NR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] \\
      (a) & (b) \\
  \end{tabular}
  \end{center}

  \caption{In this contrived and impossible position (a), white has many
    easy paths to mate.  All of black's pieces are pinned.
    There is also a ``series of legal moves''
    where black mates white: \bishop xg7++  \bishop xh8?? \Knight xe7++.
    Capturing the king (by moving twice in a row) is a ``legal move''
    (despite not being allowed by other rules), and doing so unpins
    black's knight to deliver a smothered mate (b). Of course, abuse of
    this dubious technical possibility doesn't change the status of
    the position, since we already have a mating sequence by white.
  } \label{fig:kingcapture}
\end{figure}

Second, what is a ``series'' of legal moves? It seems completely
consistent to allow the white player to make several legal moves in
a row, for example. The rules about alternating moves are
again outside the definition of ``legal move'' and ``series'' is
never defined.

We could instead interpret ``any series of legal moves'' as ``taking
the entirety of the rules of chess, any continuation of the game that
ends in checkmate for either player.'' I like this better, although it
creates its own subtle issues. For example, should the position be
considered dead if there is a checkmating sequence, but it requires
entering a fivefold repetition or exhaustion of the 75-move rule? If
so, this would end the game prematurely, and so it has implications
for the longest possible game (Section~\ref{sec:ending}). Even more
esoterically, this interpretation causes the rule to be
self-referential: A sequence must also be allowed by the rule being
defined. A normal person would take the ``least fixed point'' (in the
Kripke sense~\cite{kripke1975outline}) of this self-referential
definition (fewest positions are drawn). But it is also consistent to
interpret it {\em maximally}---in which case the longest chess game is
zero moves!

@@

It is generally not hard to avoid repeating positions, so the main
obstacle we'll face is the 75-move rule. Let's call an irreversible
move that resets the 75-move counter a {\em critical move}; this is
a pawn move or capturing move (or both). The structure of the game
will be a series of critical moves (I will call these ``critical''
moves) with a maximal sequence of pointless reversible moves in
between them. If we execute the maximum number of critical moves and
make 149 moves (just shy of triggering the compulsory version of
the 75-move rule) between them, then this will be fairly easily
seen as a maximal game.

In fact, for most positions, it is easy to waste 149 moves and return
to the exact same position. So, the strategy for generating the
longest game can mostly be broken into two tasks: Make a game with a
maximum number of critical moves (it can also contain other moves) and
then pad that game out to maximum-length excursions in between its
critical moves.\footnote{The first phase was constructed by hand.
  Software for inserting excursions and checking the result is
  available at:
  \url{sf.net/p/tom7misc/svn/HEAD/tree/trunk/chess/longest.cc} }

\subsection{Maximal critical moves} \label{sec:maxcritical}

Critical moves are pawn moves and captures. There are 16 pawns, each
of which has 6 squares to move into before promoting, so this is
$16\times 6$ critical moves. There are 14 capturable pieces, plus the
16 pawns (they can be captured after promoting); capturing them nets an
additional $16+14$ critical moves for a total of $126$. Each critical
move can be made after a maximum of $75+74$ reversible moves, giving
$(75+74+1) \times 126 = 18900$ moves. The final move would capture the
last piece, yielding a draw due to the remaining kings being insufficient
material (``dead position''). This is our starting upper bound.

@@

However, each of these captures is {\em both} a pawn move {\em and} a capturing
move. This means that we lose $4+4$ critical moves off our
total budget. $150 \times (126 - 8) = 17700$ is the new upper bound.

@@

Note that white is quite constrained during this beginning phase, as
pawn moves are critical moves and must be avoided. Only the white
knights can escape the back rank. When we try to insert $149$
pointless moves, we'll only be able to move the knights and rooks, and
doing this $75$ times must leave e.g.~one of the knights on an
opposite-colored square.\footnote{Knight moves always change the color
  of the knight's square, and same for rook moves constrained to the
  a1/b1 or g1/h1 squares.} So we have to be a little careful about the
position in which we make black's first critical move.

Since white can only free their two knights, these are the only pieces
that can be captured by black pawns. So it will not be possible for
black to double four sets of their pawns as in
Figure~\ref{fig:doubled}. This would require white to have a phase of
critical moves to free pieces to capture, then black again to finish
doubling pawns, then white again to promote its freed pawns. Each switch
costs one move off the na\"ive max. We can be more efficient with an
asymmetric approach.

@@

\comment{
1. Nf3 b6 2. Nc3 g6 3. Ne5 g5 4. Nd5 b5
5. Nc4 bxc4 6. Nf4 gxf4 7. Rg1 c6 8. Rh1 c5
9. Rg1 f6 10. Rh1 f5 
}
\begin{center}
  \chessboard[setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11,
    pgfstyle=straightmove,markmoves={a6-b7,c3-b4,d6-c7,e6-f7,f3-g4,h6-g7},
  ]
\end{center}

The white b and g pawns have a clear route to promotion. White can
free each of the remaining 6 pawns with a single capture. Between this
and black's two pawn captures, this is the optimal 8 pawn moves
that are also captures. Since black has plenty of freed pieces, white
can promote all of their pawns during their own phase of critical
moves, resulting in:

@@

As we switch which color is making critical moves a total of three
times, we must come shy of the na\"ive maximum of 17,700 moves by
three. This gives us an upper bound of 17,697 for this approach,
which we will be able to achieve.

\paragraph{Ending condition.} \label{sec:ending} The way this game ends
is subtle for several reasons. First, note that we left the white
queen on the board and used it for mate. It is required that white be
the one mating for parity reasons, similar to the reason black must
make the first critical move. Since white makes critical moves in the
last phase, black leads on non-critical moves; at the moment white
makes the checkmating move, black has made 75 non-critical moves, and
white 74. White's 75th move mates.

@@

Why not capture the white queen too? This would be a critical move,
but once two kings remain, the game ends immediately in a draw. So
there is no length advantage here over checkmate. In fact we could do
worse: Consider the position in Figure~\ref{fig:insufficient} where
black is {\em forced} to capture the queen. This game is over prior to
the capture, so this is shorter than the checkmating sequence! Even if
the king had an escape square, it is arguable
(Section~\ref{sec:ambiguity}) that the game is over in any case: There
is no series of legal moves that leads to mate. Either the king
captures the queen (and then clearly no mate is possible with just the
two kings) or the king escapes, but in doing so plays the
75\th\ non-critical move, and triggers a draw by the 75-move rule.

@@

The game described in Section~\ref{sec:maxcritical} has 118 critical
moves in 289 total moves. There is some inefficiency between the
critical moves, but this doesn't matter since we are trying to generate
a long game anyway. In fact, the next step will be to add {\em as much
  inefficiency as possible} in between the critical moves.

% Diagram here or earlier would probably help!

Each ``critical section,'' which is the series of moves ending in a
critical move, can be treated independently. If black ends the section
with a critical move, then we want $75+74$ non-critical moves to be
played, and then black's critical move. For white, of course, $74+75$.
There are many ways we could try to make these $149$ moves; we don't
even have to use the moves that are already there as long as we end
up in the right position to make the critical move. But a simple
approach suffices.

For each critical section, we loop over all of the positions
encountered, and attempt inserting excursions that return us to the
same position but waste moves. There are two types of excursions we
try: Even excursions (each player makes two moves) and odd (each
player makes three moves).

\paragraph{Even excursions.} This four move sequence moves two pieces
of $X_1$ and $X_2$ of opposite colors. $X_1$ moves from $s_1$ to $d_1$
then $X_2$ from $s_2$ to $d_2$, then $X_1$ moves from $d_1$ back to
$s_1$, and $X_2$ from $d_2$ back to $s_2$. Easy. Any piece can perform
this maneuver other than pawns (which would be critical moves anyway)
as long as there are legal squares (considering check, etc.). All of
$s_1$, $d_1$, $s_2$, $d_2$ must be distinct. No shorter excursions are
possible.

\paragraph{Odd excursions.} This is the straightforward extension to
three squares ($s_i \rightarrow m_i \rightarrow d_i \rightarrow s_i$),
for a six-move sequence. The squares for each piece must be distinct
but it is possible for e.g.~$m_2$ to equal $s_1$. Knights cannot
perform this trick; each move changes the color of the square the
knight sits on, which causes a contradiction with a cycle of length
three. All other pieces can do it with sufficient room. The king, for
example, can move horizontally, then diagonally, then vertically back
to its starting square.

Note that odd excursions are not possible early in the game (prior to
white moving any pawns), because even when the knights are free, the
rooks only have two squares (and thus the same color parity argument
applies as knights). Some opportunity can be created by having a black
knight capture one of white's bishops. Fortunately, we do not need any
odd excursions at this point in the game.

We find excursions by just looping over possible moves that satisfy
the criteria, prioritizing odd excursions if the target (divided by
two) is odd. In order to avoid triggering the fivefold repetition
rule, we also keep track of all of the positions encountered, and
never enter a position more than two times. (Here we avoid even
threefold repetition.) It is not necessary to look beyond the critical
section, because critical moves make it impossible to return to a
prior position.

This process is not at all guaranteed to work; it may fail to fill the
critical section. Indeed, as discussed in
Section~\ref{sec:maxcritical}, we must have at least one move of slack
whenever we switch from a critical move by one player to the other.
In practice this approach succeeds readily, and manages to waste 149
moves in each critical section of the input game, save for the three
times that parity requires one move of slack. The full game is uselessly
included in a very tiny font in Section~\ref{sec:longest}.

\section{Reader's guide}

The paper demonstrates a game with three ``switches'' of which side is
making critical moves; each costs a move against the na\"ive maximum
due to parity. We clearly need at least one switch (both sides must
make critical moves), but is it possible to do it with only two? If
not, can this lower bound be proved?

\smallskip
The game given is believed to be maximal, as measured in the number of
moves. But, other metrics exist. For example, the letter g is slightly
wider than f, so moving \Queen g3 is {\em typographically} longer than
\Bishop f1. PGN format itself can be stretched by making moves that
need to be disambiguated (\bishop ff7 means ``move the \bishop\ on the
f file to f7'' (not ``the giant sword from Final Fantasy 7'', as many
believe)) or checking the opponent's king for a bonus +. Some moves
are longer in terms of distance traveled; moving the queen or bishop
between opposite corners is $\sqrt{2} \times 7$ squares! What is the
longest game according to these or other metrics?

% XXX
% Are there any practical implications of the

\smallskip
In {\it Chess}, it is impossible to capture your own pieces. How does
this limitation apply to your own life?

\smallskip
In many games of {\it Chess}, the black and white pieces are found to
disagree. What does this say about society?

\smallskip
How do you feel about the ending of the game? Is it disappointing that
the result is not symmetric (i.e.~a draw), or elegant that it
demonstrates its own avoidance of the ``dead position'' rule? Was it
what you expected?

\smallskip
{\it Is this the longest Chess game?} is the author's sixth paper
about chess for SIGBOVIK, but it is widely believed that nobody wants
to read this kind of thing. What is wrong with him?

\section{A longest game} \label{sec:longest}

There are jillions of possible games that satisfy the description
above and reach 17,697 moves; here is one of them.\footnote{
  It can also be downloaded at \url{tom7.org/chess/longest.pgn}.
  Many chess programs fail to load the whole game, but this is
  because they decided not to implement the full glory of chess.}
%
Critical moves are marked with {\large \bf [bold]}. Note that in the
standard PGN format for listing games, move numbers are ``full moves''
consisting of a move by white and then black, whereas we use ``move''
in this paper to mean ``half-move''; an individual move by one of the
players. Thus the game ends during full move 8,849, one half of
17,697.

\medskip

@@
